### 2015-11-17 paper.Rmd
table and figure references

```{r points-and-lines, fig.height=6, fig.width=7, echo=FALSE, message=FALSE, fig.cap="Left side plots show an example of a survey of an area containing a population of 500 objects, blue indicates sampler placement (top lines, bottom points) and red dots indicate detected individuals. The right side of the figure shows histograms of observed distances (again, lines top and points bottom)."}
opar <- par(mfrow=c(2,2))
# line and point transect examples
# top row points/samplers in space
# bottom row histograms

@@

@@

Key function plus adjustment terms (or adjustment series) models are formulated by taking a "key" function and optionally adding "adjustments" to it to improve the fit. Mathematically we formulate this as:
$$
g(y; \boldsymbol{\theta}) = k(y; \boldsymbol{\theta}_\text{key})\left( 1+ \alpha_O(y; \boldsymbol{\theta}_\text{adjust})\right),
$$
where $k$ is the key function and $\alpha_O$ is sum series of functions (given in **Table ???**), described as an *adjustment of order $O$*. Subscripts on the parameter vector indicate those parameters belonging to each part of the model (i.e. $\boldsymbol{\theta} = (\boldsymbol{\theta}_\text{key}, \boldsymbol{\theta}_\text{adjust})$).

Some models for the key are as follows:
$$
k(y) = \left\{
\begin{array}{l l}
  \exp\left(-\frac{y^2}{2 \sigma^2}\right) & \quad \text{half-normal,} \\
  1-\exp\left(\left(-\frac{y}{\sigma}\right)^{-b}\right) & \quad \text{hazard-rate,} \\
  1/w & \quad \text{uniform.}
\end{array} \right.
$$
Possible modelling options for key and adjustments are given in Table \ref{ka-tab} and illustrated in **Figure ???**. We select the number of adjustment terms ($K$) by AIC (further details in "Model checking and model selection").

```{r hn-hr-par-comp, fig.width=10, fig.height=5, fig.cap="Half-normal (top row) and hazard-rate (bottom row) detection functions without adjustments, varying scale ($\\sigma$) and (for hazard-rate) shape ($b$) parameters (values are given above the plots). On the top row from left to right, the study species becomes more detectable (higher probability of detection at larger distances). The bottom row shows the hazard-rate model's more pronounced shoulder.", echo=FALSE}
par(mfrow=c(2,4), mar=c(3.5, 3, 2, 1) + 0.1)

@@

\begin{table}
\caption{Modelling options for key plus adjustments models for the detection function.}
\begin{tabular}{llll}
\hline
key function   & form   & adjustment & form\\
\hline
 uniform  & $1/w$   & cosine  & $\sum_{o=1}^O a_o \cos(o \pi y/w)$ \\
 & & Simple polynomial & $\sum_{o=1}^O a_o (y/w)^{2o}$ \\
 half-normal  & $\exp\left(-\frac{y^2}{2 \sigma^2}\right)$ & cosine  & $\sum_{o=2}^O a_o \cos(o \pi y/w)$ \\
 & & Hermite polynomial & $\sum_{o=2}^O a_o H_{2o}(y/\sigma)$ \\
 hazard-rate  & $1-\exp\left[-\left(\frac{y}{\sigma}\right)^{-b}\right]$ & cosine  & $\sum_{o=2}^O a_o \cos(o \pi y/w)$ \\
 & & Simple polynomial & $\sum_{o=2}^O a_o (y/w)^{2o}$ \\
\hline
\end{tabular}
\label{ka-tab}
\end{table}

@@

```{r adjust-mix-comp, fig.width=9, fig.height=3, fig.cap="Possible shapes for the detection function when adjustments are included for half-normal and hazard-rate models.", echo=FALSE}
par(mfrow=c(1,4), mar=c(3.5, 3, 2, 1) + 0.1)

@@

**Figure???** shows the result of calling \code{plot} on the resulting model object. We can also call \code{summary} on the model object to get summary information about the fitted model (though we postpone this until the next section).

@@

By default \code{ds} assumes that the data given to it is line transect, but we can switch to points using \code{transect="point"}. Including covariates in the scale is via the \code{formula=~...} argument to \code{ds}; a hazard-rate model for the amakihi that includes observer as a covariate (and truncating at 82.5m, given in @Marques:2007ey) can be specified by:
```{r amakihi-hr-obs}
amakihi_hr_obs <- ds(amakihi, truncation=82.5, transect="point",
                     key="hr", formula=~obs)
```
As with the minke whale model, we can plot the resulting detection function. **Since for the amakihi we used covariates in the detection function, the plot will show the detection function averaged over levels/values of the covariate and separately for each level set (for continuous covariates, the 25\%, 50\% and 75\% quantiles are displayed).**

@@

**When multiple covariates are used in the model, a series of plots are produced. We can see this in the bottom row of Figure ???, where we plot the results of fitting following model:**
```{r amakihi-hr-obs-mas, warning=FALSE}
amakihi_hr_obs_mas <- ds(amakihi, truncation=82.5, transect="point",
                         key="hr", formula=~obs+mas)
```

```{r minke-hn-plot, echo=FALSE, warning=FALSE, fig.width=9, fig.height=3, fig.cap="Plot of fitted detection functions overlayed on the histograms of observed distances. Left minke whale data, half-normal model; centre, amakihi data hazard-rate with observer as a covariate; right, amakihi data, hazard-rate model with observer and minutes after sunrise as covariates"}
par(mfrow=c(1,3))
plot(minke_hn)
plot(amakihi_hr_obs)
plot(amakihi_hr_obs_mas)
```

@@

We can generate a Q-Q plot and test results using the \code{gof_ddf} function. We can see the goodness of fit tests below for two models for the amakihi data, first fitting a half-normal model without covariates or adjustments (not setting \code{adjustment=NULL} will force \code{ds} to fit a model with no adjustments), then calculating goodness of fit for that model and our hazard-rate model with observer and minutes after sunrise included:
```{r amakihi-gof, fig.keep="none"}
amakihi_hr <- ds(amakihi, truncation=82.5, transect="point", key="hn", adjustment=NULL)
gof_ds(amakihi_hr)
gof_ds(amakihi_hr_obs_mas)
```
The corresponding Q-Q plots are shown in **Figure ???**.

```{r, amakihi-qq-comp, fig.width=6, fig.height=3, fig.cap="Comparison of quantile-quantile plots for a half-normal model (no adjustments, no covariates) and hazard-rate model with observer and minutes after sunrise for the amakihi data.", echo=FALSE, results="hide"}
par(mfrow=c(1,2))
gof_ds(amakihi_hr, asp=1)
gof_ds(amakihi_hr_obs_mas, asp=1)
```

@@

Once we have a set of models which fit well, we can use Akaike's Information Criterion (AIC) to select between models. \pkg{Distance} includes a function to create table of summary information for fitted models, making it easy to get an overview of a large number of models at once. The \code{summarize_ds_models} function takes models as input and can be especially useful when paired with \pkg{knitr}'s \code{kable} function to create summary tables for publication. An example of this output is shown in **Table ???** and was generated by:
```{r summary-table-null, eval=FALSE}
library(knitr)
summarize_ds_models(amakihi_hr, amakihi_hr_obs, amakihi_hr_obs_mas)
```

```{r summary-table, results="asis", echo=FALSE}
library(knitr)
kable(summarize_ds_models(amakihi_hr, amakihi_hr_obs, amakihi_hr_obs_mas),
      digits = 3, format = "latex", row.names = FALSE, escape=FALSE,
      caption="Summary for the detection function models fitted to the amakihi data. \"C-vM\" stands for Cramer-von Mises, $P_a$ indicates average detectability (see \"Estimating abundance and variance\"), se indicates standard error. Models are sorted according to AIC.")
```

@@

As an example the minke whale data consists of two strata: \code{North} and \code{South} relating to a stratum further away and nearer the Antarctic ice edge, respectively. Figure \ref{minke-strata} shows the two strata, along with observations and transect lines.

\begin{figure}
\includegraphics{minke-strata}
\label{minke-strata}
\caption{Strata used for the minke whale data adapted from @Hedley:2004et. Points show the locations of observations along transect lines. The stepped line shows the boundary between North and South strata. Further details on the survey are available in @Branch:2001ua (simulated data is based on ``1992/93 Area III'' therein).}
\end{figure}

@@

Having already fitted a model to the minke whale data, we can see the results of the abundance estimation by simply looking at a model summary:
```{r minke-N-summary}
summary(minke_hn)
```
This prints a rather large amount of information. One useful feature is that the returned object is a \code{list} of \code{data.frames}, so we can again use \pkg{knitr}s \code{kable} function to create useful tables for publication. For example **Table ???**.

```{r minke-abundance-table}
minke_table <- summary(minke_hn)$dht$individuals$N
minke_table$lcl <- minke_table$ucl <- minke_table$df <- NULL
colnames(minke_table) <- c("Stratum", "$\\hat{N}$", "$\\text{se}(\\hat{N}$)", 
                           "$\\text{CV}(\\hat{N}$)")
kable(minke_table, digits = 3, format = "latex",
      row.names = FALSE, escape=FALSE,caption="Summary of abundance estimation for the half-normal model for the minke whale data.")
```

>>>>

```{r points-and-lines, fig.height=6, fig.width=7, echo=FALSE, message=FALSE, fig.cap="Left side plots show an example of a survey of an area containing a population of 500 objects, blue indicates sampler placement (top lines, bottom points) and red dots indicate detected individuals. The right side of the figure shows histograms of observed distances (again, lines top and points bottom).\\label{fig:pointslines}"}
opar <- par(mfrow=c(2,2))
# line and point transect examples
# top row points/samplers in space
# bottom row histograms

@@

Figure \ref{fig:pointslines} shows examples of 500 individuals sampled using line and point transects (left column) and their corresponding histograms (right column).

@@

Key function plus adjustment terms (or adjustment series) models are formulated by taking a "key" function and optionally adding "adjustments" to it to improve the fit. Mathematically we formulate this as:
$$
g(y; \boldsymbol{\theta}) = k(y; \boldsymbol{\theta}_\text{key})\left( 1+ \alpha_O(y; \boldsymbol{\theta}_\text{adjust})\right),
$$
where $k$ is the key function and $\alpha_O$ is sum series of functions (given in Table \ref{tab:keyadj}), described as an *adjustment of order $O$*. Subscripts on the parameter vector indicate those parameters belonging to each part of the model (i.e. $\boldsymbol{\theta} = (\boldsymbol{\theta}_\text{key}, \boldsymbol{\theta}_\text{adjust})$).

Some models for the key are as follows:
$$
k(y) = \left\{
\begin{array}{l l}
  \exp\left(-\frac{y^2}{2 \sigma^2}\right) & \quad \text{half-normal,} \\
  1-\exp\left(\left(-\frac{y}{\sigma}\right)^{-b}\right) & \quad \text{hazard-rate,} \\
  1/w & \quad \text{uniform.}
\end{array} \right.
$$
Possible modelling options for key and adjustments are given in Table \ref{tab:keyadj} and illustrated in Figure \ref{fig:hnhr} and \ref{fig:keyadj}. We select the number of adjustment terms ($K$) by AIC (further details in "Model checking and model selection").

```{r hn-hr-par-comp, fig.width=10, fig.height=5, fig.cap="Half-normal (top row) and hazard-rate (bottom row) detection functions without adjustments, varying scale ($\\sigma$) and (for hazard-rate) shape ($b$) parameters (values are given above the plots). On the top row from left to right, the study species becomes more detectable (higher probability of detection at larger distances). The bottom row shows the hazard-rate model's more pronounced shoulder.\\label{fig:hnhr}", echo=FALSE}
par(mfrow=c(2,4), mar=c(3.5, 3, 2, 1) + 0.1)

@@

\begin{table}
\caption{Modelling options for key plus adjustment series models for the detection function.}
\begin{tabular}{llll}
\hline
key function   & form   & adjustment & form\\
\hline
 uniform  & $1/w$   & cosine  & $\sum_{o=1}^O a_o \cos(o \pi y/w)$ \\
 & & Simple polynomial & $\sum_{o=1}^O a_o (y/w)^{2o}$ \\
 half-normal  & $\exp\left(-\frac{y^2}{2 \sigma^2}\right)$ & cosine  & $\sum_{o=2}^O a_o \cos(o \pi y/w)$ \\
 & & Hermite polynomial & $\sum_{o=2}^O a_o H_{2o}(y/\sigma)$ \\
 hazard-rate  & $1-\exp\left[-\left(\frac{y}{\sigma}\right)^{-b}\right]$ & cosine  & $\sum_{o=2}^O a_o \cos(o \pi y/w)$ \\
 & & Simple polynomial & $\sum_{o=2}^O a_o (y/w)^{2o}$ \\
\hline
\end{tabular}
\label{tab:keyadj}
\end{table}

@@

```{r adjust-mix-comp, fig.width=9, fig.height=3, fig.cap="Possible shapes for the detection function when adjustments are included for half-normal and hazard-rate models.\\label{fig:keyadj}", echo=FALSE}
par(mfrow=c(1,4), mar=c(3.5, 3, 2, 1) + 0.1)

@@

Figure \ref{fig:minkeamakihi} (left panel) shows the result of calling \code{plot} on the resulting model object. We can also call \code{summary} on the model object to get summary information about the fitted model (though we postpone this until the next section).

@@

By default \code{ds} assumes that the data given to it is line transect, but we can switch to points using \code{transect="point"}. Including covariates in the scale is via the \code{formula=~...} argument to \code{ds}; a hazard-rate model for the amakihi that includes observer as a covariate (and truncating at 82.5m, given in @Marques:2007ey) can be specified by:
```{r amakihi-hr-obs}
amakihi_hr_obs <- ds(amakihi, truncation=82.5, transect="point",
                     key="hr", formula=~obs)
```
As with the minke whale model, we can plot the resulting detection function (Figure \ref{fig:minkeamakihi}, centre and right panels). *Since for the amakihi we used covariates in the detection function, the plot will show the detection function averaged over levels/values of the covariate and separately for each level set (for continuous covariates, the 25\%, 50\% and 75\% quantiles are displayed).*

@@

*When multiple covariates are used in the model, a series of plots are produced. We can see this in the bottom row of Figure \ref{fig:minkeamakihi}, where we plot the results of fitting following model:*
```{r amakihi-hr-obs-mas, warning=FALSE}
amakihi_hr_obs_mas <- ds(amakihi, truncation=82.5, transect="point",
                         key="hr", formula=~obs+mas)
```

```{r minke-amakihi-hn-plot, echo=FALSE, warning=FALSE, fig.width=9, fig.height=3, fig.cap="Plot of fitted detection functions overlayed on the histograms of observed distances. Left minke whale data, half-normal model; centre, amakihi data hazard-rate with observer as a covariate; right, amakihi data, hazard-rate model with observer and minutes after sunrise as covariates.\\label{fig:minkeamakihi}"}
par(mfrow=c(1,3))
plot(minke_hn)
plot(amakihi_hr_obs)
plot(amakihi_hr_obs_mas)
```

@@

We can generate a Q-Q plot and test results using the \code{gof_ddf} function. We can see the goodness of fit tests below for two models for the amakihi data, first fitting a half-normal model without covariates or adjustments (not setting \code{adjustment=NULL} will force \code{ds} to fit a model with no adjustments), then calculating goodness of fit for that model and our hazard-rate model with observer and minutes after sunrise included:
```{r amakihi-gof, fig.keep="none"}
amakihi_hr <- ds(amakihi, truncation=82.5, transect="point", key="hn", adjustment=NULL)
gof_ds(amakihi_hr)
gof_ds(amakihi_hr_obs_mas)
```
The corresponding Q-Q plots are shown in Figure \ref{amakihi-qq}.

```{r, amakihi-qq-comp, fig.width=6, fig.height=3, fig.cap="Comparison of quantile-quantile plots for a half-normal model (no adjustments, no covariates) and hazard-rate model with observer and minutes after sunrise for the amakihi data.\\label{amakihi-qq}", echo=FALSE, results="hide"}
par(mfrow=c(1,2))
gof_ds(amakihi_hr, asp=1)
gof_ds(amakihi_hr_obs_mas, asp=1)
```

@@

Once we have a set of models which fit well, we can use Akaike's Information Criterion (AIC) to select between models. \pkg{Distance} includes a function to create table of summary information for fitted models, making it easy to get an overview of a large number of models at once. The \code{summarize_ds_models} function takes models as input and can be especially useful when paired with \pkg{knitr}'s \code{kable} function to create summary tables for publication. An example of this output is shown in Table \ref{tab:amakihi} and was generated by:
```{r summary-table-null, eval=FALSE}
library(knitr)
summarize_ds_models(amakihi_hr, amakihi_hr_obs, amakihi_hr_obs_mas)
```

```{r summary-table, results="asis", echo=FALSE}
library(knitr)
kable(summarize_ds_models(amakihi_hr, amakihi_hr_obs, amakihi_hr_obs_mas),
      digits = 3, format = "latex", row.names = FALSE, escape=FALSE,
      caption="Summary for the detection function models fitted to the amakihi data. \"C-vM\" stands for Cramer-von Mises, $P_a$ indicates average detectability (see \"Estimating abundance and variance\"), se indicates standard error. Models are sorted according to AIC.\\label{tab:amakihi}")
```

@@

As an example the minke whale data consists of two strata: \code{North} and \code{South} relating to a stratum further away and nearer the Antarctic ice edge, respectively. Figure \ref{fig:minke-strata} shows the two strata, along with observations and transect lines.

\begin{figure}
\includegraphics{minke-strata}
\label{fig:minke-strata}
\caption{Strata used for the minke whale data adapted from @Hedley:2004et. Points show the locations of observations along transect lines. The stepped line shows the boundary between North and South strata. Further details on the survey are available in @Branch:2001ua (simulated data is based on ``1992/93 Area III'' therein).}
\end{figure}

@@

Having already fitted a model to the minke whale data, we can see the results of the abundance estimation by simply looking at a model summary:
```{r minke-N-summary}
summary(minke_hn)
```
This prints a rather large amount of information. One useful feature is that the returned object is a \code{list} of \code{data.frames}, so we can again use \pkg{knitr}s \code{kable} function to create useful tables for publication. For example Table \ref{minke-abund}.

```{r minke-abundance-table}
minke_table <- summary(minke_hn)$dht$individuals$N
minke_table$lcl <- minke_table$ucl <- minke_table$df <- NULL
colnames(minke_table) <- c("Stratum", "$\\hat{N}$", "$\\text{se}(\\hat{N}$)", 
                           "$\\text{CV}(\\hat{N}$)")
kable(minke_table, digits = 3, format = "latex",
      row.names = FALSE, escape=FALSE,caption="Summary of abundance estimation for the half-normal model for the minke whale data.\\label{minke-abund}")
```
